Modeling and analyzing the Corona-virus warning app with the Isabelle Infrastructure framework

### 2022-02-28 DPM2020/final/paper.tex
moved the folder final into DPM2020 as its relevant there

The app now in use in Germany has been developed based on a protocol proposed by the DP-3T project \cite{dp3t:github}:
A sophisticated security concept
conceived by experts in the field that has strong claims with regard to mathematical support \cite{dp3t:wp} [p2].
%
However, there has, as of yet, to our knowledge no formal verification been involved.
Even if a ``post-production'' formal specification seems pointless, 
it allows to reveal weak points of the architecture, show that the measures that have
been conceived are suitable to cover those weak points,
or to what extent trade offs have to be made due to inherent vulnerabilities.
The protocol implemented by the framework CWA resembles only the most basic
of the three protocols proposed by the DP-3T project.
Hence, a formal verified model that stresses the impact or limits of certain security measures
might give more weight to appeals like \cite{dp3t:impl}
to adopt the more sophisticated protocols.

The contributions of this paper are (a) formal re-engineering of CWA
(b) providing an additional security and privacy analysis with interactive theorem proving
certification of a novel view on the system architecture including actors, locations, and policies,
(c) a formal definition of a security refinement process that allows to improve a system
based on attacks found by the attack tree analysis and (d) an application of the refinement to
improve security of the CWA specification.

@@

The formal model in the Isabelle insider framework is fully mechanized and proved in
Isabelle (sources available \cite{kam:18smc}).

@@

In terms of attack scenarios, we focus on, what might be classified as deanonymisation attacks: Tracking a device (see \cite{dp3t:psre}[p9], \cite{pepppt:dp3tana}[p8]) and identifying infected individuals (\cite{dp3t:psre}[p5], \cite{pepppt:dp3tana}[p9]).

@@

\subsubsection{Kripke structures, CTL, and Attack Trees}
\label{sec:intro}
The Isabelle framework has now after various case studies become a general framework for
the state-based security analysis of infrastructures with policies and actors. Temporal logic
and Kripke structures build the foundation. 
Meta-theoretical results have been established to show equivalence between attack trees 
and CTL statements \cite{kam:18b}. 
%As part of the meta-theory Correctness and Completeness have been 
%proved in Isabelle \cite{kam:18b} and can be used to navigate between CTL and attack 
%trees to find attacks. 
This foundation provides a generic notion of state transition on which attack trees and
temporal logic can be used to express properties. 
The main notions used in this paper are:
\begin{itemize}
\item {Kripke structures and state transitions:}\\ 
Using a generic state transition relation $\mapsto$, Kripke structures
are defined as a set of states reachable by $\mapsto$ from an initial
state set, for example
\begin{ttbox}
Kripke \{t. \ttexists i \ttin I. i \ttrelIstar t\} I
\end{ttbox}
\item {CTL statements:}\\ 
For example, we can write 
\begin{ttbox}
K \ttvdash {\sf EF} s
\end{ttbox}
to express that in Kripke structure \texttt{K} there is a path on which
the property \texttt{s} (a set of states) will eventually hold.
\item {Attack trees:} \\
The datatype of attack trees has three constructors: 
$\oplus_\vee$ creates or-trees and $\oplus_\wedge$ creates 
and-trees.
And-attack trees $l \ttattand s$ and or-attack trees $l \ttattor s$ 
consist of a list of sub-attacks -- again attack trees. The third constructor 
creates a base attack as a pair of state sets written \texttt{\ttcalN{(I,s)}}.
For example, a two step and-attack leading from state set \texttt{I} via
\texttt{si} to \texttt{s} is expressed as
\begin{ttbox}
\ttvdash [\ttcalN{(I,si)},\ttcalN{(si,s)}]\ttattand{\texttt{(I,s)}}
\end{ttbox}
%\begin{ttbox}
%\ttvdash [\ttcalN{(I,s)},\ttcalN{(s,s')}]\ttattand{(I,s')}
%\end{ttbox}
\item {Attack tree refinement, validity and adequacy:}\\
Attack trees have their own refinement (not to be mixed up with the
system refinement presented in this paper as introduced in the next section).
An abstract attack tree may be refined by spelling out the attack steps until a valid attack
is reached:

\texttt{\ttvdash A :: (\ttsigma :: state) attree)}.

The validity is defined constructively (code is generated from it)
and its adequacy with respect to a formal semantics in CTL is proved
and can be used to facilitate actual application verification as demonstrated 
her in the stepwise system refinements.
\end{itemize}

\subsubsection{Instantiation of Framework}
The formal model of CWA uses the Isabelle Infrastructure framework instantiating it 
by reusing its concept of {\it actors} for users and smartphones whereby locations correspond
to physical locations. The Ephemeral Ids, their sending and change is added to Infrastructures
by slightly adapting the basic state type of infrastructure graphs and accordingly the semantic rules
for the actions move, get, and put. The details of the newly adapted Infrastructure are
presented in Section \ref{sec:model}.
Technically, an Isabelle theory file \texttt{Infrastructure.thy} builds on top of the theories for Kripke 
structures and CTL (\texttt{MC.thy}), attack trees (\texttt{AT.thy}), and security refinement 
(\texttt{Refinement.thy}). Thus all these concepts can be used to specify the formal model
for CWA, express relevant and interesting properties, and conduct
interactive proofs (with the full support of the powerful and highly automated proof support
of Isabelle).
%The CWA theory itself is an adaptation of the Infrastructure theory of the 
%Isabelle Infrastructure framework and reuses (or slightly adapts) existing concepts. 
%In the remainder of this paper, we introduce the model that we conceived for . 
All Isabelle sources are available online \cite{kam:20gitsc}.

\subsubsection{Refinement}
An additional feature that has been integrated into the Isabelle Infrastructure
framework motivated by security engineering formal specifications for IoT healthcare
system is an extension of the formal specification process introducing
refinement of Kripke structures \cite{kam:19a,kam:20a}. It refines a system model based on a 
formal definition of a combination of trace refinement and structural 
refinement (or datatype refinement). The definition allows to prove property preservation results 
crucial for an iterative development process.
The refinements of the system specification  can be interleaved with attack 
analysis while security properties can be proved in Isabelle. In each iteration
security qualities are accumulated while continuously attack trees scrutinize
the design.
One of the contributions of this paper is to explore different concepts of refinement:
the formal expression of refinement, enables to pin down (i.e. exemplify) different concepts
of refinement (data refinement, action refinement, trace refinement (aka spec refinement) and
combinations thereof with concrete attack scenarios.

@@

The transition between states is triggered by non-parameterized
actions \texttt{get}, \texttt{move}, and \texttt{put} 
executed by actors. 
Actors are given by an abstract type \texttt{actor} and a function 
\texttt{Actor} that creates elements of that type from identities 
(of type \texttt{string} written \texttt{''s''} in Isabelle). 
Actors are contained in an infrastructure graph type \texttt{igraph}
constructed by its constructor \texttt{Lgraph}. 
\begin{ttbox}
{\bf datatype} igraph = 
         Lgraph (location \tttimes location)set 
                 location \ttfun identity set
                 identity \ttfun (string set \tttimes string set \tttimes efid)  
                 location \ttfun string \tttimes (dlm \tttimes data) set
                 location \ttfun efid set
                 actor \ttfun location \ttfun (identity \tttimes efid) set
\end{ttbox}
In the current application of the framework to the CWA case study, 
this graph contains a set of location pairs representing the %structure 
topology of the infrastructure
as a graph of nodes and a function%\footnote{ We use the common $\lambda$-abstraction, 
%e.g. \texttt{\ttlam x. True}, to define functions with parameters, here the 
%function returning True for any input $x$.} 
that assigns a set of actor identities to each node (location) in the graph.
The third component of an \texttt{igraph} assigns the credentials to each actor: 
a triple-valued function whose first range component is a set describing the credentials 
in the possession of an actor and the second component 
is a set defining the roles the actor can take on; 
most prominently the third component is the \texttt{efid} assigned to the actor. This is initially
just a natural number but will be refined to actually represent lists of Ephemeral Ids later
when refining the specification.
\begin{ttbox}
{\bf datatype} efid = Efid nat
\end{ttbox}
The fourth component of the type \texttt{igraph}
assigns security labeled data to locations, a feature not used in the
current application.
The second to last component assigns the set of efids of all currently present smart phones
to each location of the graph. The last component finally denotes the knowledge set of each
actor for each location: a set of pairs of actors and potential ids.

@@

The Privacy protection goal is to avoid deanonymization. That is, an attacker should not be able to
disambiguate the set of pairs of real ids and EphIDs. This is abstractly expressed in the predicate
identifiable.
\begin{ttbox}
identifiable eid A \ttequiv is_singleton\{(Id,Eid). (Id, Eid) \ttin A \ttand Eid = eid\}
\end{ttbox}
The predicate identifiable is used to express as the global policy `Eve cannot deanonymize an Ephemeral
Id \texttt{eid} using the gathered knowledge':
%{\bf fixes} global_policy::[infrastructure, identity] \ttfun bool
%{\bf defines}  
\begin{ttbox}
  global_policy I eid \ttequiv
             \ttneg(identifiable eid 
                ((\ttbigcap (kgra(graphI I)(Actor ''Eve'')`(nodes(graphI I))))
                 - \{(x,y). x = ''Eve''\}))
\end{ttbox}

\subsection{Infrastructure state transition}
\label{sec:statetrans}
The state transition relation uses the syntactic infix notation 
\texttt{I \ttrelI\, I'}  to denote that infrastructures 
\texttt{I} and \texttt{I'} are in this relation.
%Rules for the put, get, move, process and delete actions exist.
To give an impression of this definition, we show first the rule defining the state transition for
the action get. Initially, this rule expresses that
an actor that resides at a location \texttt{l} (\texttt{a \ttatI\ l})
and is enabled by the local policy in this location to ``get'' can
combine  all ids at the current location (contained in \texttt{egra G l}) with
all actors at the current location (contained in \texttt{agra G l}) and add this
set of pairs to his knowledge set \texttt{kgra G} using the function update \texttt{f(l := n)}
redefining the function \texttt{f} for the input \texttt{l} to have now the new value \texttt{n}.
\begin{ttbox}
 {\bf{get\_data}}: G = graphI I \ttImp a \ttatI l \ttImp enables I l (Actor a) get \ttImp 
   I' = Infrastructure 
          (Lgraph (gra G)(agra G)(cgra G)(lgra G)(egra G)
            ((kgra G)((Actor a) := ((kgra G (Actor a))(l:=
                 \{(x,y). x \ttin agra G l \ttand y \ttin egra G l\})))))
          (delta I)
   \ttImp I \ttrel{n} I' 
\end{ttbox}
Another interesting rule for the state transition is the one for \texttt{move} whose structure
resembles the previous one.
\begin{ttbox}
 {\bf{move}}: G = graphI I \ttImp a \ttatI l \ttImp a \ttin actors_graph(graphI I) \ttImp
    l \ttin nodes G \ttImp l' \ttin nodes G \ttImp enables I l' (Actor h) move \ttImp
    I' = Infrastructure (move_graph_a a l l' (graphI I))(delta I)
 \ttImp I \ttrelI I' 
\end{ttbox}
The semantics of this rule is embedded in the function \texttt{move\_graph\_a}
that adapts the infrastructure state so that the moving actor \texttt{a} is now
associated to the target location \texttt{l'} in the actor map \texttt{agra} and not
any more at \texttt{l} and also the association of \texttt{efids} is updated accordingly.
\begin{ttbox}
move_graph_a n l l' g \ttequiv
  Lgraph (gra g) 
         (if n \ttin ((agra g) l) \ttand  n \ttnin ((agra g) l') then 
           ((agra g)(l := (agra g l) - \{n\}))(l' := (insert n (agra g l')))
          else (agra g))
         (cgra g)(lgra g)
         (if n \ttin ((agra g) l) \ttand  n \ttnin ((agra g) l') then
               ((egra g)(l := (egra g l) - \{efemid (cgra g n)\}))
                        (l' := (insert (efemid (cgra g n))(egra g l')))
          else egra g)
         (kgra g)
\end{ttbox}  
Based on this state transition and the above defined \texttt{Corona\_scenario}, we define the first Kripke structure.
\begin{ttbox}
 corona_Kripke \ttequiv Kripke \{ I. Corona_scenario \ttrelIstar I \} \{Corona_scenario\}
\end{ttbox}

@@

The attack tree calculus \cite{kam:18b} exhibits that an attack is possible.
\begin{ttbox}
 corona_Kripke \ttvdash {\sf EF} scorona
\end{ttbox}
We can simply apply the Correctness theorem \texttt{AT\_EF} to 
immediately prove this CTL statement. This application of the meta-theorem 
of Correctness of attack trees saves us proving the CTL formula tediously 
by exploring the state space in Isabelle proofs. Alternatively, we could use 
the generated code for the function \texttt{is\_attack\_tree} in Scala 
(see \cite{kam:18b}) to check that a refined attack of the above is valid.

\section{Refinement}
\label{sec:ref}
%\TODO{This section is just copied over from arxive paper - adapt}
Intuitively, a refinement changes some aspect of the type of
the state, for example, replaces a data type by a richer datatype or
restricts the behaviour of the actors. The former is expressed directly 
by a mapping of datatypes, the latter is incorporated into the state
transition relation of the Kripke structure that corresponds to the 
transformed model.
%This relationship between the Kripke structure can be better understood
%visually (see Figure \ref{fig:modtrans})
In other words, we can encode a refinement within our framework
as a relation on Kripke structures that is parameterized additionally by
a function that maps the refined type to the abstract type.
The direction ``from refined to abstract'' of this type mapping may seem 
curiously counter-intuitive. However, the actual refinement is given by the 
refinement that uses this function as an input. The refinement
then refines an abstract to a more concrete system specification. 
The additional layer for the refinement 
%(the red box in Figure \ref{fig:theorystruc})
can be formalized in Isabelle as a
%second\footnote{The first refinement 
%relation in this framework is on attack trees summarized in Section \ref{sec:at}.} 
refinement relation 
$\sqsubseteq_{\mathcal{E}}$. 
The relation \texttt{mod\_trans} is typed as a relation over triples --
a function from a threefold Cartesian product to \texttt{bool}, the 
type containing true and false only.  
The type variables $\sigma$ and $\sigma'$ input to the type constructor 
\texttt{Kripke} represent the abstract state type and the concrete state type. 
Consequently, the middle element of the triples selected by the relation 
\texttt{mod\_trans} is a function of type $\sigma' \Rightarrow \sigma$ 
mapping elements of the refined state to the abstract state.
The expression in quotation marks after the type is again the
infix syntax in Isabelle that allows the definition of mathematical notation
instead of writing \texttt{mod\_trans} in prefix manner. This nicer infix
syntax is already used in the actual definition.
Finally, the arrow \texttt{\ttImp} is the implication of Isabelle's  meta-logic
while $\ttimp$ is the one of the {\it object} logic HOL. 
They are logically equivalent but of different
types: within a HOL formula $P$, for example, as below $\forall x. P \ttimp Q$, only the implication $\ttimp$
can be used.
\begin{ttbox}
 refinement :: (\ttsigma Kripke \tttimes (\ttsigma' \ttfun \ttsigma) \tttimes \ttsigma' Kripke) \ttfun bool ("_ \ttmref{(\_)} _")
  K \ttmeref K' \ttequiv \ttforall s' \ttin states K'. \ttforall s \ttin init K'. 
             s \ttrelstar{\sigma'} s' \ttimp \ttecal(s) \ttin init K \ttand \ttecal(s) \ttrelstar{\sigma} \ttecal(s')
\end{ttbox}
The definition of \texttt{K \ttmeref\, K'} states that for any state $s'$ 
of the refined Kripke structure that can be reached by the state transition
in zero or more steps from an initial state $s$ of the refined Kripke 
structure, the mapping ${\mathcal E}$ from the refined to the abstract 
model's state must preserve this reachability, i.e., the image of
$s$ must also be an initial state and from there the image of $s'$
under ${\mathcal E}$ must be reached with $0$ or $n$ steps.

@@

represents function image, that is the set, $\{\ttecal(x). x \in \texttt{init K'}\} $.

@@

structure refinement entails property preservation and makes it possible 

@@

allows to explicitly formalize such sophisticated concepts like reachability.

@@

For cases, where we want to have the simpler one-step proviso but still need 
reachability we provide a slightly weaker version of \texttt{strong\_mt}.

@@

the properties of a system are given by the set of its traces. Now, a refinement
of the system is given by another system that has a subset of the traces of the 

@@

additionally incorporates refinement. Since we include a state map 

@@

From a cryptographic point of view, the crucial properties of the Ephemeral Ids are that
they are purely random, therefore, they cannot be guessed, but at the same time if -- after
the actual encounter between sender and receiver -- the seed $SK_t$ is published, it is feasible
to relate any of the \texttt{EphIDi} to $SK_t$  for all \texttt{i} $\in \{1, \ldots, n\}$.
For a formalization of this crucial cryptographic property in Isabelle it suffices to define a new type
of list of Ephemeral Ids \texttt{efidlist} containing the root $SK_t$ (the first \texttt{efid}),
a current \texttt{efid} indicated by a list pointer of type \texttt{nat}, and the actual list of
\texttt{efid}s.
\begin{ttbox}
{\bf{datatype}} efidlist = Efids "efid" "nat" "efid list"
\end{ttbox}
We define functions for this datatype: \texttt{efidsroot} returning the first of the three
constituents in an \texttt{efidlist} (the root $SK_t$); \texttt{efids\_index} giving the second
component, the index of the current \texttt{efid}; \texttt{efids\_inc\_ind} applied to an
\texttt{efidlist} increments the index; \texttt{efids\_cur} returning the current \texttt{efid}
from the list and \texttt{efids\_list} for the entire list (the third component).

>>>>

The app now in use in Germany has been developed based on a protocol proposed by the DP-3T project \cite{dp3t:github}:
A sophisticated security concept
conceived by experts in the field that has strong claims with regard to mathematical support \cite{dp3t:wp} [p2].
%
However, there has, as of yet, to our knowledge no formal verification been involved.
Even if a ``post-production'' formal specification seems pointless, 
it allows revealing weak points of the architecture, show that the measures that have
been conceived are suitable to cover those weak points,
or to what extent trade-offs have to be made due to inherent vulnerabilities.
The protocol implemented by the framework CWA resembles only the most basic
of the three protocols proposed by the DP-3T project.
Hence, a formal verified model that stresses the impact or limits of certain security measures
might give more weight to appeals like \cite{dp3t:impl}
to adopt the more sophisticated protocols.

The contributions of this paper are (a) formal re-engineering of CWA
(b) providing an additional security and privacy analysis with interactive theorem proving
certification of a novel view on the system architecture including actors, locations, and policies,
(c) a formal definition of a security refinement process that allows improving a system
based on attacks found by the attack tree analysis and (d) an application of the refinement to
improve security of the CWA specification.

@@

The formal model in the Isabelle insider framework is fully mechanized and proved in
Isabelle (sources available \cite{kam:20gitat}).

@@

In terms of attack scenarios, we focus on, what might be classified as deanonymization attacks: Tracking a device (see \cite{dp3t:psre}[p9], \cite{pepppt:dp3tana}[p8]) and identifying infected individuals (\cite{dp3t:psre}[p5], \cite{pepppt:dp3tana}[p9]).

@@

\subsubsection{Kripke structures, CTL, and Attack Trees}
\label{sec:intro}
The Isabelle framework has now after various case studies become a general framework for
the state-based security analysis of infrastructures with policies and actors. Temporal logic
and Kripke structures build the foundation. 
Meta-theoretical results have been established to show equivalence between attack trees 
and CTL statements \cite{kam:18b}. 
%As part of the meta-theory Correctness and Completeness have been 
%proved in Isabelle \cite{kam:18b} and can be used to navigate between CTL and attack 
%trees to find attacks. 
This foundation provides a generic notion of state transition on which attack trees and
temporal logic can be used to express properties. 
The main notions used in this paper are:
\begin{itemize}
\item {Kripke structures and state transitions:}\\ 
Using a generic state transition relation $\to$, Kripke structures
are defined as a set of states reachable by $\to$ from an initial
state set, for example,
\begin{ttbox}
Kripke \{t. \ttexists i \ttin I. i \ttrelIstar t\} I
\end{ttbox}
where $\to^*$ is the reflexive transitive closure over $\to$.
\item {CTL statements:}\\ 
For example, we can write 
\begin{ttbox}
K \ttvdash {\sf EF} s
\end{ttbox}
to express that in Kripke structure \texttt{K} there is a path on which
the property \texttt{s} (a set of states) will eventually hold.
\item {Attack trees:} \\
The datatype of attack trees has three constructors: 
$\oplus_\vee$ creates or-trees and $\oplus_\wedge$ creates 
and-trees.
And-attack trees $l \ttattand s$ and or-attack trees $l \ttattor s$ 
consist of a list of sub-attacks -- again attack trees. The third constructor 
creates a base attack as a pair of state sets written \texttt{\ttcalN{(I,s)}}.
For example, a two step and-attack leading from state set \texttt{I} via
\texttt{si} to \texttt{s} is expressed as
\begin{ttbox}
\ttvdash [\ttcalN{(I,si)},\ttcalN{(si,s)}]\ttattand{\texttt{(I,s)}}
\end{ttbox}
%\begin{ttbox}
%\ttvdash [\ttcalN{(I,s)},\ttcalN{(s,s')}]\ttattand{(I,s')}
%\end{ttbox}
\item {Attack tree refinement, validity and adequacy:}\\
Attack trees have their own refinement (not to be mixed up with the
system refinement presented in this paper as introduced in the next section).
An abstract attack tree \texttt{A} of type \texttt{\ttalpha\ attree} over an arbitrary state type \texttt{\ttsigma}
may be refined by spelling out the attack steps until a valid attack is reached:

\texttt{\ttvdash A :: (\ttsigma :: state) attree}.

The validity is defined constructively (code is generated from it)
and its adequacy with respect to a formal semantics in CTL is proved
and can be used to facilitate actual application verification as demonstrated 
here in the stepwise system refinements.
\end{itemize}

\subsubsection{Instantiation of Framework}
The formal model of CWA uses the Isabelle Infrastructure framework instantiating it 
by reusing its concept of {\it actors} for users and smartphones whereby locations correspond
to physical locations. The Ephemeral Ids, their sending and change is added to infrastructures
by slightly adapting the basic state type of infrastructure graphs and accordingly the semantic rules
for the actions move, get, and put. The details of the newly adapted infrastructure are
presented in Section \ref{sec:model}.
Technically, an Isabelle theory file \texttt{Infrastructure.thy} builds on top of the theories for Kripke 
structures and CTL (\texttt{MC.thy}), attack trees (\texttt{AT.thy}), and security refinement 
(\texttt{Refinement.thy}). Thus all these concepts can be used to specify the formal model
for CWA, express relevant and interesting properties, and conduct
interactive proofs (with the full support of the powerful and highly automated proof support
of Isabelle).
%The CWA theory itself is an adaptation of the Infrastructure theory of the 
%Isabelle Infrastructure framework and reuses (or slightly adapts) existing concepts. 
%In the remainder of this paper, we introduce the model that we conceived for . 
All Isabelle sources are available online \cite{kam:20gitat}.

\subsubsection{Refinement}
An additional feature that has been integrated into the Isabelle Infrastructure
framework motivated by security engineering formal specifications for IoT healthcare
system is an extension of the formal specification process introducing
refinement of Kripke structures \cite{kam:19a,kam:20a}. It refines a system model based on a 
formal definition of a combination of trace refinement and structural 
refinement (or datatype refinement). The definition allows proving property preservation results 
crucial for an iterative development process.
The refinements of the system specification  can be interleaved with attack 
analysis while security properties can be proved in Isabelle. In each iteration
security qualities are accumulated while continuously attack trees scrutinize
the design.
One of the contributions of this paper is to explore different concepts of refinement:
the formal expression of refinement, enables to pin down (i.e. exemplify) different concepts
of refinement (data refinement, action refinement, trace refinement (aka spec refinement) and
combinations thereof with concrete attack scenarios.

@@

The transition between states is triggered by non-parameterized
actions \texttt{get}, \texttt{move}, and \texttt{put} 
executed by actors. 
Actors are given by an abstract type \texttt{actor} and a function 
\texttt{Actor} that creates elements of that type from identities 
(of type \texttt{string} written \texttt{''s''} in Isabelle). 
Actors are contained in an infrastructure graph type \texttt{igraph}
constructed by its constructor \texttt{Lgraph}. 
\begin{ttbox}
{\bf datatype} igraph = 
         Lgraph (location \tttimes location)set 
                 location \ttfun identity set
                 identity \ttfun (string set \tttimes string set \tttimes efid)  
                 location \ttfun string \tttimes (dlm \tttimes data) set
                 location \ttfun efid set
                 actor \ttfun location \ttfun (identity \tttimes efid) set
\end{ttbox}
In the current application of the framework to the CWA case study, 
this graph contains a set of location pairs representing the %structure 
topology of the infrastructure
as a graph of nodes and a function %\footnote{ We use the common $\lambda$-abstraction, 
%e.g. \texttt{\ttlam x. True}, to define functions with parameters, here the 
%function returning True for any input $x$.} 
that assigns a set of actor identities to each node (location) in the graph.
The third component of an \texttt{igraph} assigns the credentials to each actor: 
a triple-valued function whose first range component is a set describing the credentials 
in the possession of an actor and the second component 
is a set defining the roles the actor can take on; 
most prominently the third component is the \texttt{efid} assigned to the actor. This is initially
just a natural number of type \texttt{nat} mapped by the constructor \texttt{Efid} to the new datatype
\texttt{efid} but will be refined to actually represent lists of Ephemeral Ids later
when refining the specification.
\begin{ttbox}
{\bf datatype} efid = Efid nat
\end{ttbox}
The fourth component of the type \texttt{igraph}
assigns security labeled data to locations, a feature not used in the
current application.
The second to last component assigns the set of efids of all currently present smart phones
to each location of the graph. The last component finally denotes the knowledge set of each
actor for each location: a set of pairs of actors and potential ids.

@@

The Privacy protection goal is to avoid deanonymization. That is, an attacker should not be able to
disambiguate the set of pairs of real ids and EphIDs. This is abstractly expressed in the predicate
\texttt{identifiable}.
\begin{ttbox}
identifiable eid A \ttequiv is_singleton\{(Id,Eid). (Id, Eid) \ttin A \ttand Eid = eid\}
\end{ttbox}
The predicate \texttt{identifiable} is used to express as the global policy `Eve cannot deanonymize an Ephemeral
Id \texttt{eid} using the gathered knowledge':
%{\bf fixes} global_policy::[infrastructure, identity] \ttfun bool
%{\bf defines}  
\begin{ttbox}
  global_policy I eid \ttequiv
             \ttneg(identifiable eid 
                ((\ttbigcap (kgra(graphI I)(Actor ''Eve'')`(nodes(graphI I))))
                 - \{(x,y). x = ''Eve''\}))
\end{ttbox}

\subsection{Infrastructure state transition}
\label{sec:statetrans}
The state transition relation uses the syntactic infix notation \texttt{I \ttrel\, I'}  to denote that
infrastructures \texttt{I} and \texttt{I'} are in this relation. 
We will use this state transition to define the dynamic behaviour of the system including actors' movements and
the evolution of their knowledge. This allows representing attacks as sequences of state transitions leading to
states in which the global policy is violated. 
To give an intuitive motivating idea before the formal presentation in Section \ref{sec:ana}: the attack will
consist of the state transitions of Eve getting available EphIds at the location pub, Alice moving to shop, Eve
following her and again getting the EphIds at shop.

So, one relevant rule for the state transition is the one for \texttt{move} that defines the state transition relation
expressing state changes when actors move across the infrastructure.
%whose structure resembles the previous one.
\begin{ttbox}
 {\bf{move}}: G = graphI I \ttImp a \ttatI l \ttImp a \ttin actors_graph(graphI I) \ttImp
    l \ttin nodes G \ttImp l' \ttin nodes G \ttImp enables I l' (Actor h) move \ttImp
    I' = Infrastructure (move_graph_a a l l' (graphI I))(delta I)
 \ttImp I \ttrel I' 
\end{ttbox}
The semantics of this rule is embedded in the function \texttt{move\_graph\_a}
that adapts the infrastructure state so that the moving actor \texttt{a} is now
associated to the target location \texttt{l'} in the actor map \texttt{agra} and not
any more at \texttt{l} and also the association of \texttt{efids} is updated accordingly.
\begin{ttbox}
move_graph_a n l l' g \ttequiv
  Lgraph (gra g) 
         (if n \ttin ((agra g) l) \ttand  n \ttnin ((agra g) l') then 
           ((agra g)(l := (agra g l) - \{n\}))(l' := (insert n (agra g l')))
          else (agra g))
         (cgra g)(lgra g)
         (if n \ttin ((agra g) l) \ttand  n \ttnin ((agra g) l') then
               ((egra g)(l := (egra g l) - \{efemid (cgra g n)\}))
                        (l' := (insert (efemid (cgra g n))(egra g l')))
          else egra g)
         (kgra g)
\end{ttbox}  
%Rules for the put, get, move, process and delete actions exist.
%To give an impression of this definition,
Another action relevant for the deanonymization attack describes how an actor gets information.
This is defined by the rule of the state transition for the action get. Initially, this rule
expresses that an actor that resides at a location \texttt{l} (\texttt{a \ttatI\ l})
and is enabled by the local policy in this location to ``get'' can
combine  all ids at the current location (contained in \texttt{egra G l}) with
all actors at the current location (contained in \texttt{agra G l}) and add this
set of pairs to his knowledge set \texttt{kgra G} using the function update \texttt{f(l := n)}
redefining the function \texttt{f} for the input \texttt{l} to have now the new value \texttt{n}.
\begin{ttbox}
 {\bf{get\_data}}: G = graphI I \ttImp a \ttatI l \ttImp enables I l (Actor a) get \ttImp 
   I' = Infrastructure 
          (Lgraph (gra G)(agra G)(cgra G)(lgra G)(egra G)
            ((kgra G)((Actor a) := ((kgra G (Actor a))(l:=
                 \{(x,y). x \ttin agra G l \ttand y \ttin egra G l\})))))
          (delta I)
   \ttImp I \ttrel I' 
\end{ttbox}
Based on this state transition and the above defined \texttt{Corona\_scenario}, we define the first Kripke structure.
\begin{ttbox}
 corona_Kripke \ttequiv Kripke \{ I. Corona_scenario \ttrelIstar I \} \{Corona_scenario\}
\end{ttbox}

@@

The attack tree calculus \cite{kam:18b} exhibits that an attack is possible.
\begin{ttbox}
 corona_Kripke \ttvdash {\sf EF} scorona
\end{ttbox}
We can simply apply the Correctness theorem \texttt{AT\_EF} \cite{kam:18b,kam:18a}
\begin{ttbox}
\ttvdash A \ttImp attack A = (I,s)
\ttImp Kripke \{s . \ttexists i \ttin I. (i \ttrelIstar s) \} I \ttvdash EF s
\end{ttbox}  
to immediately prove this CTL statement. This application of the meta-theorem 
of Correctness of attack trees saves us proving the CTL formula tediously 
by exploring the state space in Isabelle proofs. Alternatively, we could use 
the generated code for the function \texttt{is\_attack\_tree} in Scala 
(see \cite{kam:18b}) to check that a refined attack of the above is valid.

\section{Refinement}
\label{sec:ref}
Intuitively, a refinement changes some aspect of the type of
the state, for example, replaces a data type by a richer datatype or
restricts the behaviour of the actors. The former is expressed directly 
by a mapping of datatypes, the latter is incorporated into the state
transition relation of the Kripke structure that corresponds to the 
transformed model.
%This relationship between the Kripke structure can be better understood
%visually (see Figure \ref{fig:modtrans})
In other words, we can encode a refinement within our framework
as a relation on Kripke structures that is parameterized additionally by
a function that maps the refined type to the abstract type.
The direction ``from refined to abstract'' of this type mapping may seem 
curiously counter-intuitive. However, the actual refinement is given by the 
refinement that uses this function as an input. The refinement
then refines an abstract to a more concrete system specification. 
The additional layer for the refinement 
%(the red box in Figure \ref{fig:theorystruc})
can be formalized in Isabelle as a
%second\footnote{The first refinement 
%relation in this framework is on attack trees summarized in Section \ref{sec:at}.} 
refinement relation 
$\sqsubseteq_{\mathcal{E}}$. 
This relation, called \texttt{refinement}, is typed as a relation over triples --
a function from a threefold Cartesian product to \texttt{bool}, the 
type containing true and false only.  
The type variables $\sigma$ and $\sigma'$ input to the type constructor 
\texttt{Kripke} represent the abstract state type and the concrete state type. 
Consequently, the middle element of the triples selected by the relation 
\texttt{refinement} is a function of type $\sigma' \Rightarrow \sigma$ 
mapping elements of the refined state to the abstract state.
The expression in quotation marks after the type is again the
infix syntax in Isabelle that allows the definition of mathematical notation
instead of writing \texttt{refinement} in prefix manner. This nicer infix
syntax is already used in the actual definition.
Finally, the arrow \texttt{\ttImp} is the implication of Isabelle's  meta-logic
while $\ttimp$ is the one of the {\it object} logic HOL. 
They are logically equivalent but of different
types: within a HOL formula, for example, within the below $\forall $\texttt{s' \ttin\ states K'.}\dots,
only the implication $\ttimp$ can be used.
\begin{ttbox}
 refinement :: (\ttsigma Kripke \tttimes (\ttsigma' \ttfun \ttsigma) \tttimes \ttsigma' Kripke) \ttfun bool ("_ \ttmref{(\_)} _")
  K \ttmeref K' \ttequiv \ttforall s' \ttin states K'. \ttforall s \ttin init K'. 
             s \ttrelstar{\sigma'} s' \ttimp \ttecal(s) \ttin init K \ttand \ttecal(s) \ttrelstar{\sigma} \ttecal(s')
\end{ttbox}
The definition of \texttt{K \ttmeref\, K'} states that for any state $s'$ 
of the refined Kripke structure that can be reached by the state transition
in zero or more steps from an initial state $s$ of the refined Kripke 
structure, the mapping ${\mathcal E}$ from the refined to the abstract 
model's state must preserve this reachability, i.e., the image of
$s$ must also be an initial state and from there the image of $s'$
under ${\mathcal E}$ must be reached within $0$ and $n$ steps.

@@

represents function image, i.e.,  $\{\ttecal(x) \mid x \in \texttt{init K'}\} $.

@@

structures entails property preservation and makes it possible 

@@

allows explicitly formalizing such sophisticated concepts like reachability.

@@

For cases, where we %want to
have the simpler one-step proviso %but still need 
and also reachability from initial states 
we provide a slightly weaker version of \texttt{strong\_mt}.
%that pre-supposes reachability of \texttt{s} from initial states.

@@

the properties of a system are given by the set of its event traces. Now, a refinement
of the system is given by another system that has a subset of the event traces of the 

@@

additionally incorporates data refinement. Since we include a state map 

@@

From a cryptographic point of view, the crucial properties of the Ephemeral Ids are that
they are purely random, therefore, they cannot be guessed, but, at the same time, if -- after
the actual encounter between sender and receiver -- the seed $SK_t$ is published, it is feasible
to relate any of the \texttt{EphIDi} to $SK_t$  for all \texttt{i} $\in \{1, \ldots, n\}$.
For a formalization of this crucial cryptographic property in Isabelle it suffices to define a new type
of list of Ephemeral Ids \texttt{efidlist} containing the root $SK_t$ (the first \texttt{efid}),
a current \texttt{efid} indicated by a list pointer of type \texttt{nat}, and the actual list of
\texttt{efid}s.
\begin{ttbox}
{\bf{datatype}} efidlist = Efids "efid" "nat" "efid list"
\end{ttbox}
We define functions for this datatype: \texttt{efidsroot} returning the first of the three
constituents in an \texttt{efidlist} (the root $SK_t$ of type \texttt{efid}); \texttt{efids\_index} giving the second
component of type \texttt{nat}, that is, the index of the current \texttt{efid}; \texttt{efids\_inc\_ind} applied to an
\texttt{efidlist} increments the index; \texttt{efids\_cur} returning the current \texttt{efid}
from the list and \texttt{efids\_list} for the entire list (the third component of type \texttt{efid list}).
